Some lessons after the formalization of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-05-30 Independence_CH/lessons.tex
Misprints after second read

We believe that in every project of formalization of mathematics,
there is a tension between the haste to verify the target results and
the need to obtain a readable, albeit extremely detailed, corpus of
statements and proofs. This tension is mirrored in two different
purposes of formalization: Developing new mathematics from scratch and
producing verified results en route to this, versus verifying and
documenting material that has already been produced on paper.

@@

A similar experience, but on an opposite side of the formalization
spectrum happened to the Liquid Tensor Project as described by Scholze
in \cite{LTE2021}. People involved in the formalization simply pushed
their way to reach the summit, formalizing lemma after lemma. They
actually wrote the blueprint for that formalization \emph{afterwards}
it was complete! From time to time, we were also frenziedly trying to
get the results formalized, going beyond what we had planned.

@@

\subsection{How to believe in the formalization}
This is a rather tricky question, that was addressed by Pollack in his
\cite{MR1686867}. There is little point to discuss that, after an
assistant has accepted some input successfully, \emph{some
mathematics} has been formally verified. What might not be apparent is
if the claimed theorems are indeed the results that have been
checked. One key aspect of this is the logical foundation of the
assistant (Section~\ref{sec:isabelle-metalogic-meta}). But the weakest
link in the chain is the laying down of definitions building up to the
concepts needed to state the target results. (Another related aspect,
concerning the way results are printed and parsed by assistant versus
there internal meaning, was studied by Wiedijk \cite{zbMATH06319597}.)

@@

Frequently, we formalized material by directly typing the proof we
new by heart, and in doing so we assumed that some definitions
accommodated some of our preconceptions.
It is significant that in a few such occasions, we were doubly
surprised by the fact that some supposedly trivial lemma would not go
through, because the definitions addressed something different (think
of \emph{restriction} of a function to a set versus that of a
relation), and also that we were able to proved the previous
results. The takeaway is that intuition may drive proofs
even if you are not working on what we believe we are.

@@

Some of those proofs were almost copy-pasted once and again with minor
variants; this would usually be relegated to some function in the
meta-language, but we were unable to do this due to our limitations in
programming Isabelle/ML\footnote{%
  On the other hand, our inability to automate proofs of replacement
  instances paved the way for identifying which were the ones needed for
  forcing! ($\leftarrow$\textbf{put somewhere, don't know if this is the right
    place}).}
and  (an ML style having several quirks and which
is not very well document).

Nevertheless, the experience of the CS members of the team (and
community) is invaluable for the kind of engineering that is needed in
a large project like the present one. For example, it is
(mathematically) misleading when automatic tools (\isatt{simp},
\isatt{auto}, etc) stop working just because of the sheer size of the
goal (v.g., the same statement with 7 variables succeed but with 8
variables doesn't). Scale issues are very easily disregarded in the
abstract; for example, the formula $\forceisa({\cdot}0\in 1{\cdot})$
can be explicitly printed by Isabelle (it spans nearly 20k symbols), but
$\forceisa({\cdot}\neg{\cdot}\neg{\cdot}0\in 1{\cdot}{\cdot}{\cdot})$
can not.

Another example concerns the very definition of $\forceisa$. As a
proof of concept, one of us tried to obtain its definition by
exclusively using formula synthesis, which was supposed to be trivial,
since it is mathematically (as in the case of
Equation~(\ref{eq:sats_big_union_fm})).
But in fact, some early minor mistake
rendered the whole effort useless. We then turned to a more
informed programming discipline, which involved decomposing the
definition in stages, each of which was checked for correctness, and
in that way we were able to reach our objective.

>>>>

We believe that in every project of formalization of mathematics,
there is a tension between the haste to verify the target results and
the need to obtain a readable, albeit extremely detailed, corpus of
statements and proofs. This tension is mirrored in two different
purposes of formalization: Developing new mathematics from scratch and
producing verified results on the way, versus verifying and
documenting material that has already been produced on paper.

@@

A similar experience, but on an opposite side of the formalization
spectrum happened to the Liquid Tensor Experiment as described by Scholze
in \cite{LTE2021}. People involved in the formalization simply pushed
their way to reach the summit, formalizing lemma after lemma. They
actually wrote the blueprint for that formalization \emph{afterwards}
it was complete! From time to time, we were also frenziedly trying to
get the results formalized, going beyond what we had planned.

@@

\subsection{How to believe in the formalization}
This is a rather tricky question, that was addressed by Pollack in his
\cite{MR1686867}. There is little point to discuss that, after an
assistant has accepted some input successfully, \emph{some
mathematics} has been formally verified. What might not be apparent is
if the claimed theorems are indeed the results that have been
checked. One key aspect of this is the logical foundation of the
assistant (Section~\ref{sec:isabelle-metalogic-meta}). But the weakest
link in the chain is the laying down of definitions building up to the
concepts needed to state the target results. (Another related aspect,
concerning the way results are printed and parsed by assistant versus
their internal meaning, was studied by Wiedijk \cite{zbMATH06319597}.)

@@

Frequently, we formalized material by directly typing the proof we
knew by heart, and in doing so we assumed that some definitions
accommodated some of our preconceptions.
It is significant that in a few such occasions, we were doubly
surprised by the fact that some supposedly trivial lemma would not go
through, because the definitions addressed something different (think
of \emph{restriction} of a function to a set versus that of a
relation), and also that we were able to prove the adjacent
results. The takeaway is that intuition may drive proofs
even if you are not working on what you believe we are.

@@

Some of those proofs were almost copy-pasted once and again with minor
variants; this would usually be relegated to some function in the
meta-language, but we were unable to do this due to our limitations in
programming Isabelle/ML\footnote{%
  On the other hand, our inability to automate proofs of replacement
  instances paved the way for identifying which were the ones needed for
  forcing! ($\leftarrow$\textbf{put somewhere, don't know if this is the right
    place}).}
(an ML style having several quirks and which
is not very well document).

Nevertheless, the experience of the CS members of the team (and
community) is invaluable for the kind of engineering that is needed in
a large project like the present one. For example, it is
(mathematically) misleading when automatic tools (\isatt{simp},
\isatt{auto}, etc) stop working just because of the sheer size of the
goal (v.g., the same statement with 7 variables succeed but with 8
variables doesn't). Scale issues are very easily disregarded in the
abstract but, as an example, the formula $\forceisa({\cdot}0\in 1{\cdot})$
can be explicitly printed by Isabelle (it spans nearly 20k symbols), but
$\forceisa({\cdot}\neg{\cdot}\neg{\cdot}0\in 1{\cdot}{\cdot}{\cdot})$
can not.

Another point where CS expertise was a prime asset was
the very definition of $\forceisa$. As a
proof of concept, one of us tried to obtain its definition by
using formula synthesis exclusively, which was supposed to be trivial,
since it is mathematically (as in the case of
Equation~(\ref{eq:sats_big_union_fm})).
But in fact, some early minor mistake
rendered the whole effort useless. We then turned to a more
informed programming discipline, which involved decomposing the
definition in stages, each of which was checked for correctness, and
in that way we were able to reach our objective.
